---
title: "chapter2"
author: "RK"
date: "05 November 2018"
output: html_document
---

```{r data wrangling, include=FALSE, echo=FALSE}
# DATA WRANGLING
library(dplyr)

# Access the gglot2 library
library(ggplot2)

library(GGally)

dat <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-data.txt", header = TRUE, sep = "\t")

str(dat)
dim(dat)
View(dat)

## The table has 183 rows and 60 columns.

dat$attitude <- dat$Attitude / 10

#Create an analysis dataset with the variables gender, age, attitude, deep, stra, surf and points

# questions related to deep, surface and strategic learning

deep_questions <- c("D03", "D11", "D19", "D27", "D07", "D14", "D22", "D30", "D06",  "D15", "D23", "D31")

surface_questions <- c("SU02","SU10","SU18","SU26", "SU05","SU13","SU21","SU29","SU08","SU16","SU24","SU32")

strategic_questions <- c("ST01","ST09","ST17","ST25","ST04","ST12","ST20","ST28")

# select the columns related to deep learning and create column 'deep' by averaging

deep_columns <- select(dat, one_of(deep_questions))

dat$deep <- rowMeans(deep_columns)

# select the columns related to surface learning and create column 'surf' by averaging

surface_columns <- select(dat, one_of(surface_questions))

dat$surf <- rowMeans(surface_columns)

# select the columns related to strategic learning and create column 'stra' by averaging

strategic_columns <- select(dat, one_of(strategic_questions))

dat$stra <- rowMeans(strategic_columns)

# choose a handful of columns to keep

keep_columns <- c("gender", "Age", "attitude", "deep", "stra", "surf", "Points")

# select the 'keep_columns' to create a new dataset

learning2014 <- select(dat, one_of(keep_columns))

# see the structure of the new dataset

str(learning2014)

# change some column names

colnames(learning2014)[2] <- "age"

colnames(learning2014)[7] <- "points"

colnames(learning2014)

# select rows where points is greater than zero
learning2014 <- filter(learning2014, points > 0)
str(learning2014)

# create dataset

# set working directory
setwd("C:/Users/RoopeOK/Documents/Yliopisto/IODS-project")

# write csv
write.csv(learning2014, file = "learning2014.csv")

#Reading learning2014.csv

read.csv(file = "learning2014.csv")

# ~~~~~~~~~ DATA WRANGLING OVER ~~~~~~~~
```

## Chapter 2, Data Analysis


This week I did some data wrangling on the JYTOPKYS2 dataset (and created my own learning2014.csv file) and subsequently ran some descriptive analyses as well as multiple and simple regression analyses on the data. You'll find the code for the wrangling in the rmd file. The conclusion of these analyses was that, according to this dataset, study attitudes affect the statistics course scores (points) significantly (p < 0.001).


*Step 1:*

I inspected the structure and dimensions of the learning2014 dataset I created in the data wrangling process.

The data had 166 observations and 7 variables. The variables are: gender (of student), age (of student), attitude (general attitude towards statistics), deep (deep learning strategies), stra (strategic learning strategies), surf (surface learning strategies) and points (score of statistics course).

*Step 2:*

Next I drew made some descriptive tables and a scatterplot matrix to explore a (graphical) overview of the data:

```{r data analysis, include=TRUE, echo=FALSE}
# DATA ANALYSIS:

str(learning2014)
dim(learning2014)

# Learning2014 has 166 rows and 7 columns. Like it should. Yay.

# Graphical overview of the data
# Strongest correlations are between points and attitude.

p <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

p

```

Here are some observations:
Age: Age is (as expected) heavily positively skewed (since the participants are students). Mean age is 25.5. Age does not correlate strongly with any of the other variables.  
Attitude: Attitude seems to be quite normally distributed, with slight negative skew. Attitude correlates moderately with points (r = 0.437). This is an interesting finding.  
Deep: Deep learning is quite neatly normally distributed. Deep learning has weak negative correlation (r = -0.324) with surface learning. This makes sense: when deep learning increases in a student, we can expect surface learning to decrease.  
Stra: Strategic learning has a nice normal distribution, but correlations with the other variables are at best very weak.  
Surf: Surface learning correlates very weakly with the other variables (nothing significant here) except for deep learning (See above).  
Points: (When greater than zero) points averaged (mean) at 22.7, with the lowest score being 7 and the highest 33. The points distribution is negatively skewed. The distribution peaks around it's mean, but also has a smaller peak around very high scores (around 30 points). Points have a moderate correlation with the attitude varialbe (r = 0.437).

*Step 3:*

Then I did a multiple regression model; lets see how attitude, stra and surf affect points. 

```{r regression, include=TRUE, echo=FALSE}

multiple_model <- lm(points ~ attitude + stra + surf, data = learning2014)
multiple_model
summary(multiple_model)

```

Here is a summary for this analysis:

With the other variables constant, as attitudes increase with one (on a scale 0-5) points increase by ~3.4 (regression coefficient is 3.39). This is a highly significant relation (p < 0.001). The other variables (stra and surf) were not statistically significant (with respective p.values of 0.12 and 0.46 and regression coefficients of 0.85 and -0.57).

Multiple R-squared is 0.21. This implies that our model explains 21 % of the variance in points.

Since stra and surf were not significant, I chose to delete them form my model. This left me with a (nice) simple regression model with points as the dependent variable and attitude as the independent/explanatory variable:

```{r regression2, include=TRUE, echo=FALSE}
p1 <- ggplot(learning2014, aes(x = attitude, y = points)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Student's attitude versus exam points")
p1

single_model <- lm(points ~ attitude, data = learning2014)
summary(single_model)
```

*Step 4:*

The effect of attitude on points is highly significant. Multiple R-squared is 0.19, implying that the variable attitude alone accounts for 19 % of the variance in student scores. F-statistic is 14.13 with 162 degrees of freedom.

Of course... What this does NOT imply is that attitudes cause high scores. There might be a third variable lurking behind attitudes towards statistics, such as general intelligence :-)

*Step 5:*
And then for some diagnostics:


```{r diagnostics, include=TRUE, echo=FALSE}
par(mfrow = c(2,2))

plot(single_model, which = c(1, 2, 5))
```

We use the QQ plot here to evaluate how well the distribution of errors in our dataset matches a standard normal distribution. The Q-Q plot above suggests that our dataset indeed is normally distributed. Only very slight symptoms of kurtosis and skew are shown. The Residuals vs Fitted plot is nicely evenly distributed with no significant patterns standing out, suggesting that the assumption of constant variance of errors is reasonable. Although we have some outlier cases (data points 145, 56 and 35), according to our Residuals vs Leverage plot these don’t have high leverage on our dataset, since the leverage values are relatively low. Conclusively, it’s safe to say that our model fulfils the assumptions required for a linear regression analysis (i.e. our findings are reliable).

That's all for this week, folks.
